Spatiotemporal Multi-Resolution Approximation of the Amari Type Neural Field Model

### 2012-06-17 Multiresolution_framework/ltx/manuscript.tex
Paper submitted to NeuroImage.

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{setspace}                               
\usepackage[usenames,dvipsnames]{color}  
%This bit is how to use hyperlink without  coloured rectangles around it
\usepackage{float}
\usepackage{hyperref}
\hypersetup{colorlinks,linkcolor=black,filecolor=black,urlcolor=black,citecolor=black}
\usepackage{psfrag}
\newcommand{\dean}[1]{\textcolor{green}{#1}}
\newcommand{\parham}[1]{\textcolor{blue}{#1}}
\newcommand{\vik}[1]{\textcolor{red}{#1}}

%\newfloat{algorithm}{htbp}{loa}%[chapter]
%\floatname{algorithm}{Algorithm}
\journal{Neuroimage}
\begin{document}
\begin{frontmatter}
\title{Spatiotemporal Multi-Resolution Approximation of the Amari Type Neural Field Model}

@@

\begin{abstract}
%\boldmath
Neural fields are spatially continuous state variables described by integro-differential equations, which are well suited to describe the spatiotemporal evolution of cortical activations on multiple scales. Here we develop a multi-resolution approximation (MRA) framework for the integro-difference equation (IDE) neural field model based on semi-orthogonal cardinal B-spline wavelets. In this way, a flexible framework is created, whereby both macroscopic and microscopic behavior of the system can be represented simultaneously. State and parameter estimation is performed using the Expectation Maximization (EM) algorithm. A synthetic example is provided
to demonstrate the framework.
\end{abstract} 
\begin{keyword}
  Neural Field Model, multi-resolution approximation (MRA), Expectation Maximization (EM) algorithm
\end{keyword}

\end{frontmatter}

\section{Introduction}
The human cerebral cortex has a multi-resolution architecture, where spatial scales for information processing range from ion channels, to single neurons, to networks of millions of neurons. This multi-resolution cortical architecture poses major modeling challenges to efficiently describe the brain's dynamics. This paper introduces a multi-resolution data-driven framework for neural field modeling, the Multi-Resolution Approximate Integro-Difference Equation (MRAIDE) to address this challenge.

@@

The utility of these models for answering questions in clinical neuroscience and neurology lies in the ability to infer physiological parameters from electrophysiological data. The ability to create patient-specific neural models will contribute to our knowledge base of diseases such as epilepsy and will enable the development of new treatment strategies. Large-scale neuroinformatic efforts, such as the Virtual Brain project (\href{http://www.thevirtualbrain.org}{www.thevirtualbrain.org} \citep{Jirsa2010}), are on the way to enable patient-specific full brain modeling. This is particularly relevant to the advent of new devices that use therapeutic electrical stimulation. Stimulation strategies for devices currently operate in an open loop, where stimulation parameters are chosen using a process of trial and error. Therefore, there exists enormous potential to improve the performance of these devices using systems theory, which in turn requires a suitable dynamic model. Neural mass and neural field models are ideal candidates for estimation algorithms due to their strong links with physiology and their parsimony. It is expected that parameters of the neural fields models will be patient-specific and will, therefore, need to be inferred from data.

@@

\section{Multi-Resolution Cortical Architecture}
The human neocortex has an inherent multi-resolution architecture. The spatiotemporal neural field can be written as 
\begin{equation}
	v\left(\mathbf{r},t\right) = \sum_{j} v_j\left(\mathbf{r},t\right),
\end{equation}
where $v\left(\mathbf{r},t\right)$ is a voltage field that may be measured with a multi-electrode array or voltage sensitive dye, $\mathbf{r}\in \mathbb{R}^n$ (where $n\le3$) describes the spatial location within the field, $t$ denotes time, and $j$ indexes the different levels of resolution in the field. This multi-resolution structure is depicted in Fig.~\ref{fig:MultiLayerFieldModel}.
\begin{figure}[t]
	% \centering
		\includegraphics[scale=1]{./Graph/fig1.pdf}
	\caption{{\bf Multi-layer neural field structure}. The inset figures show the shape of the
connectivity kernel components at each spatial resolution.}
	\label{fig:MultiLayerFieldModel}
\end{figure}

@@

Typically, the experimentalist only has access to the net field generated by the multi-resolution architecture of the cortex. Following this, it is natural to ask how do we best describe the multi-resolution structure of the neural system. The solution to this problem was proposed by \citet{Breakspear2005} in using a wavelet decomposition. This paper is directed at developing a framework for patient-specific multi-resolution neural field models, forming a fundamental link between data and theory.

\section{IDE Neural Field Model}
\singlespacing
\begin{table*}[!t]
\begin{tabular}{|l|l|l|}
	\hline
	\textbf{Symbol} & \textbf{Quantity} & \textbf{Units}\\
	\hline
	\multicolumn{3}{|c|}{\emph{Domain and indices}}\\
	\hline
	$\Omega$ & Spatial domain & n.a.\\
	$\mathbf{r}$ & Spatial location & [mm, mm]\\
	$t$ & Time & s\\
	\hline
	\multicolumn{3}{|c|}{\emph{Model}}\\
	\hline
    $y_t$ & Observation & mV\\
    $v(\mathbf{r},t)$ & Mean membrane potential & mV \\
	$f(v\left(\mathbf{r},t\right))$ & Firing rate function & spikes.s$^{-1}$\\
	$e_t(\mathbf{r})$ & Field disturbance, with covariance function $\eta(\mathbf r-\mathbf r')$ & mV\\
	$\boldsymbol\epsilon_t$ & Observation noise, with covariance matrix $\Sigma_\epsilon$ & mV\\
	$m(\mathbf{r}_n)$ & Observation kernel where $n$ is sensor index $n=1,..,n_y$ & n.a. \\
	$w(\mathbf{r})$ & Spatial connectivity kernel & n.a.\\
	$h(t)$ & Post-synaptic response kernel & n.a.\\
	$\tau$, $\xi$ & Membrane time constant \& time constant parameter & s, n.a.\\
	$\delta(t)$ & Dirac-delta function & n.a.\\
	\hline    
	\multicolumn{3}{|c|}{\emph{Multi-resolution Approximation}} \\
	\hline                                                   
	$\phi_{j,l}(r)$&Scaling function&n.a\\
	$\psi_{j,l}(r)$&Wavelet function&n.a\\  
	$\alpha_{j_0,l}, \beta_{j,l}$&Connectivity kernel approximation and detail coefficients&mV~spike$^{-1}$\\ 
	$x_{t,j_{0},l},\check{x}_{t,j,l}$&Field approximation and detail coefficients at a given time&mV~spike$^{-1}$\\ 
	$N_m(r)$&$m^{th}$ order B-spline function&n.a.\\
	$\varphi_m(r)$&$m^{th}$ order B-spline wavelet function&n.a.\\
	\hline
	\multicolumn{3}{|c|}{\emph{State-Space Model}} \\
	\hline
		$\boldsymbol\mu(r)$&Vector of $n_x$ field scaling and wavelet functions&n.a.\\
		$\boldsymbol\lambda(r)$&Vector of $n_{\boldsymbol\theta}$ connectivity kernel scaling and wavelet functions&n.a.\\
   	$\mathbf{x}_t$ & State vector at time $t$ composed of  $x_{t,j_{0},l}$ and $\check{x}_{t,j,l}$ & mV\\ 
		$\boldsymbol\theta$&Vector of connectivity kernel parameters composed of $\alpha_{j_0,l}$ and $\beta_{j,l}$& mV.spike$^{-1}$\\ 
		$\boldsymbol{\Lambda}_x$&Inner product of field basis functions&n.a.\\
		$\boldsymbol\Gamma, \boldsymbol\Lambda_{\boldsymbol{\theta}}$&Decomposition connectivity matrices&s, mV.spike$^{-1}$\\
   	$\mathbf{w}_t$ & State disturbance vector, with covariance $\Sigma_w$ & mV\\ 
    $\mathbf{A}(\boldsymbol{\theta})$& State transition matrix& n.a.\\
   	$\mathbf{C}$ & Observation matrix & n.a. \\
	\hline
	\multicolumn{3}{|c|}{\emph{Frequency Analysis}} \\
	\hline
	$\boldsymbol{\nu}$& spatial frequency  & cycles/mm \\
	\hline
	\multicolumn{3}{|c|}{\emph{Estimation}} \\
	\hline
	$\hat{\mathbf{x}}_t^{f-}$, $\hat{\mathbf{x}}_t^f$ & Forward prior and posterior state estimates & mV\\
	$\hat{\mathbf{x}}_t^{b}$ & Backward posterior state estimates & mV\\
	$\mathbf P^{f-}_t$, $\mathbf P^f_t$  & Forward prior and posterior covariance matrices & mV$^2$\\
	$\mathbf P^b_t$ & Backward posterior covariance matrices & mV$^2$\\
	$\mathbf M_t$& Cross-covariance matrix & mV$^2$\\
	$\mathcal K_{t} $ & Filter gain & n.a.\\ 
	$\mathcal S_{t} $ & Smoother gain & n.a.\\ 
	$\boldsymbol\Xi_1, \boldsymbol\Xi_0$&Expected covariance and cross-covariance matrices& mV$^2$\\ 
	$\mathcal{Q}(\boldsymbol{\theta},\boldsymbol\theta')$&Lower bound on log-likelihood function&n.a.\\
	$\boldsymbol\nu_0$, $\boldsymbol\nu_1$&Lower bound coefficient matrices&n.a.\\ 
	$\boldsymbol\Upsilon$&Lower bound coefficient matrix&mV$^{-1}$.spike\\
	\hline
\end{tabular}
\caption{\textbf{Notation.} The first column of the table gives the symbols for all the notation used in the paper. The second column provides a brief description of the quantity, and the third column provides the SI units where relevant.}
\label{tab:Notation}
\end{table*}  
\doublespacing

In a small network with no time delays the single layer homogeneous neural field model, in the most general form, is described by the equation
\begin{equation}
	\label{FullDoubleIntModel} \underbrace{v\left(\mathbf{r},t\right)}_{\text{field}} =
	\int_{-\infty}^t 
	\underbrace{h\left(t - t'\right)}_{\text{synapse}} \int_\Omega
	\theta\underbrace{w\left(\mathbf{r},\mathbf{r}'\right)}_{\text{connectivity}} 
	\underbrace{f\left( v\left( \mathbf{r}',t'\right)\right)}_{\text{firing}}
	\, \mathrm{d}\mathbf{r}'\mathrm{d}t',
\end{equation}
where the post-synaptic membrane voltage at time $t$ of a population of neurons at position $\mathbf r$ within the spatial domain $\Omega$ is denoted by $v\left(\mathbf r,t\right)$. The function $h(t)$ describes the post-synaptic response kernel, $w\left(\mathbf{r},\mathbf{r}'\right)$ describes the connectivity kernel and $f(v(\mathbf{r},t))$ describes the relationship between the mean membrane potential and the mean firing rate. The parameter $\theta$ can be thought of as the synaptic efficiency or coupling strength, where $\theta>0$ and $\theta<0$ describe excitation and inhibition, respectively. There are various forms of $h(t)$ that have been proposed in the literature, with different levels of complexity. These are shown in Fig.~\ref{fig:Modelcomponents}. Similarly, both the connectivity kernel and the activity function (firing rate) can take various forms. These are also shown in Fig.~\ref{fig:Modelcomponents}. The form of $w(\mathbf{r},\mathbf{r}')$ typically describes a Laplacian (exponential) or Gaussian decay in connectivity strength from any point in the field \cite{Breakspear2010,Atay2005}. By assuming the form of $h(t)$ is the same across layers (scales) the Amari style neural field model is formed, incorporating a multilayer structure into the connectivity kernel. In this way, the connectivity kernel typically takes the form of a Mexican hat (Gaussian shape within layer connectivity) or Wizard hat (Laplacian shape within layer) as depicted in Fig.~\ref{fig:Modelcomponents}.

The stochastic IDE form of the Amari neural field formulation~\cite{Amari1977} is given by (see~\cite{Freestone2011} for a full derivation)
\begin{equation}\label{eq:DiscreteTimeModel}
	v_{t+1}\left(\mathbf{r}\right) = 
	\xi v_t\left(\mathbf{r}\right) + 
	T_s \int_\Omega { 
	    w\left(\mathbf{r},\mathbf{r'}\right)
	    f\left(v_t\left(\mathbf{r}'\right)\right) 
	\, \mathrm{d}\mathbf{r}'} 
	+ e_t\left(\mathbf{r}\right), 
\end{equation}
where in this first-order (in time) formulation of the model the membrane dynamics are included through the parameter $\xi=1-T_s/\tau$, $\tau$ is the membrane time constant and $T_s$ is the sampling time step. 
% The connectivity strength between neural populations at a distance $\mid\mathbf{r}-\mathbf{r'}\mid$ is described by the connectivity kernel $w\left(\mathbf{r}-\mathbf{r}'\right)$. The connectivity kernel is taken as a ``Mexican hat'' function, which describes local excitation and lateral inhibition \cite{Amari1977,Atay2005}. The components of the model are shown in Fig.~\ref{fig:Modelcomponents}
\begin{figure}[!t]
\centering
\includegraphics[scale=1]{./Graph/fig2.pdf}
\caption{ {\bf The IDE neural field model components}. (\textbf{A}) The Mexican hat (solid line) and Wizard hat (dotted line) connectivity kernels composed of Gaussian and Laplacian basis functions respectively. (\textbf{B}) First order (dotted line) and second order (solid line) synaptic kernels. (\textbf{C}) The sigmoidal (solid line) and the Heaviside (dotted line) activation functions.}

\label{fig:Modelcomponents}
\end{figure}
The term $e_t(\mathbf r)$ is a zero mean Gaussian disturbance, spatially colored but temporally white, with covariance function 
\begin{equation}
cov\left(e_{t}\left(\mathbf{r}\right),e_{t+t'}\left(\mathbf{r'}\right)\right)=\sigma_d\eta(\mathbf{r}-\mathbf{r'})\delta(t-t'),
\label{eq:FieldDisturbance}
\end{equation}
where $\eta(\mathbf{r}-\mathbf{r'})$ is a spatially homogeneous covariance function, $\sigma_d$ is disturbance variance and $\delta(\cdot)$ is the Dirac-delta function. %$ = \varsigma v_t(\mathbf{r})$ \cite{VanRotterdam1982,Murphy2009}.

The observation equation describing the intracranial electrophysiological recordings is given by
\begin{equation}\label{eq:ObservationEquation}
	y_t(\mathbf{r}_{n_y}) = \int_{\Omega} { m\left(\mathbf{r}_{n_y}-\mathbf{r}'\right) v_t\left(\mathbf{r}'\right) \, \mathrm{d}\mathbf{r}'} + \epsilon_t(\mathbf{r}_{n_y}), 
\end{equation}
where $m\left(\mathbf{r}_{n_y}-\mathbf{r}'\right)$ is the observation kernel at location $\mathbf{r}_{n_y}$ and  $\boldsymbol{\epsilon}_{t}\sim \mathcal{N}\left(\mathbf{0},\mathbf{\Sigma}_{\epsilon}\right)$  is an i.i.d. Gaussian white noise process and the covariance matrix $\mathbf{\Sigma}_{\epsilon}=\sigma^2_{\epsilon}\mathbf I_{n_y}$, where $\mathbf I$ denotes the identity matrix. The observation kernel models the pickup range of the electrical fields of the brain. The lead field parameters are not included in the observation equation as the intracranial recordings are considered here.

\section{MRA of the IDE in State-Space}
The purpose of forming the multi-resolution approximation (MRA) of the integro-difference equation (IDE) is to form a multi-resolution finite-dimensional state-space model that best captures the spatiotemporal characteristics of the neural field. Given such a model, it is then possible to perform state and parameter estimation in an efficient and robust manner, thus allowing for subject-specific models. The goal of this section is to demonstrate how to formulate a state-space model in a canonical form such that standard algorithms may be applied to estimate the multi-resolution connectivity structure.   
% form a parameterized equation for coefficients of the field basis functions $\mathbf{x}_t$, which becomes the state vector.

@@

% At each spatial scale indexed by $j$ (level of decomposition), the neural field can be written as
% \begin{equation}
% 	v_{j,t}\left(\mathbf{r}\right) = \begin{cases}
% 	\sum_{l \in \mathbb{Z}}x_{j,l}(t) \underbrace{\phi_{j,l}\left(\mathbf{r}\right)}_{\text{scaling function}} + \check{x}_{t,j,l}\underbrace{\psi_{j,l}\left(\mathbf{r}\right)}_{\text{mother wavelet}} & \text{if $j=0$},\\
% 	\sum_{l \in \mathbb{Z}}\check{x}_{t,j,l}\psi_{j,l}\left(\mathbf{r}\right) & \text{otherwise}
% \end{cases}
% \end{equation}
The dynamic neural field is decomposed using
\begin{equation}
 v_t\left(r\right)=\sum_{l \in \mathbb{Z}}x_{t,j_{0},l} \underbrace{\phi_{j_0,l}\left(r\right)}_{\text{scaling functions}} + \sum_{j\geq j_0}^{\infty} \sum_{l \in \mathbb{Z}} \check{x}_{t,j,l} \underbrace{\psi_{j,l}\left(r\right)}_{\text{wavelets}},
\label{eq:FieldExpansion}
\end{equation}
where $x_{t,j_{0},l}$ and $\check{x}_{t,j,l} $ are the dynamic coefficients of the expansion, constituting the state vector at time $t$. The decomposition enables a separation of the spatial characteristics, which are considered static functions that are weighted by dynamic coefficients.

@@

% where the unknown parameter and state vectors, $\boldsymbol\theta \in \mathbb{R}^{n_{\theta}}$ and $\mathbf{x}_t \in \mathbb{R}^{n_x}$, \dean{($\boldsymbol\theta \in \mathbb{R}^{n_{\theta}\times (J+2)}$ and $\mathbf{x}_t \in \mathbb{R}^{n_x \times (J+2)}$)} are defined as 
% \begin{align}
% \boldsymbol\theta^\top \triangleq& [\begin{array}{ccccc} \boldsymbol\alpha_{j_0}^\top & \boldsymbol\beta_{j_0}^\top & \boldsymbol\beta_{j_0+1}^\top & \cdots & \boldsymbol\beta_{J}^\top \end{array}] 
% \label{KernelWeights} \\
% \mathbf{x}_{t}^\top \triangleq& [\begin{array}{ccccc}\mathbf{x}_{t,j_{0}}^\top &  \check{\mathbf{x}}_{t,j_{0}}^\top & \check{\mathbf{x}}_{t,j_{0}+1}^\top & \cdots & \check{\mathbf{x}}_{t,J}^\top\end{array}].
% \label{FieldWeights}
% \end{align}
% The kernel and field approximation coefficient vectors, $\boldsymbol \alpha_{j_0}$ and $\mathbf{x}_{t,j_{0}}$, contain all the coefficients $\left\lbrace\alpha_{j_0, \mathbf{l}}:\mathbf{l} \in \mathbb{Z}^n \right\rbrace $ and $\left\lbrace x_{t,j_0, \mathbf{l}}: \mathbf{l} \in \mathbb{Z}^n\right\rbrace$, respectively. Similarly, the kernel and the field detail coefficient vectors, $\boldsymbol\beta_{j}$ and $\check{\mathbf{x}}_{t,j}$, contain all the coefficients $\left\lbrace \beta_{j,\mathbf{l}} :\mathbf{l} \in \mathbb{Z}^n\right\rbrace$ and $\left\lbrace  \check x_{t,j,\mathbf{l}}:\mathbf{l} \in \mathbb{Z}^n\right\rbrace$, respectively.
% 
% The vectors of the kernel and the field scaling and wavelet functions, $\boldsymbol\lambda$ and $\boldsymbol\mu$, respectively, are defined by
% \begin{align}
%     \label{KernelBasisVector}
%     \boldsymbol\lambda^\top(\mathbf{r}) \triangleq& \left[
%     \begin{array}{ccccc} \boldsymbol\phi_{j_0}^\top(\mathbf{r}) &
%     \boldsymbol\psi_{j_0}^\top(\mathbf{r}) & 
%     \boldsymbol\psi_{j_0+1}^\top(\mathbf{r}) &
%     \cdots &
%     \boldsymbol\psi_{J}^\top(\mathbf{r})\end{array}\right] \\
%     \label{FieldBasisVector}
%     \boldsymbol\mu^\top (\mathbf{r}) \triangleq& \left[
%     \begin{array}{ccccc}\boldsymbol\phi_{j_0}^\top(\mathbf{r}) &
%     \boldsymbol\psi_{j_0}^\top(\mathbf{r}) & 
%     \boldsymbol\psi_{j_0+1}^\top(\mathbf{r}) &
%     \cdots &
%     \boldsymbol\psi_{J}^\top(\mathbf{r})\end{array}\right]
% \end{align}
% 
% The vectors in  (\ref{KernelBasisVector}) and (\ref{FieldBasisVector}) are constructed in the same manner as the vectors in  (\ref{KernelWeights}) and (\ref{FieldWeights}).  

@@

The state-transition equation can be simplified if the chosen family of scaling and wavelet functions are isotropic where
	\boldsymbol{\lambda}_{i}(r-r') = \boldsymbol{\lambda}(2\mathbf{c}_{i}+r'-r), 
where $c_i$ is the centre of the $i^{th}$ basis function of the kernel decomposition. Following this, we can switch the order of operations with the inner product and convolution in Eq.~\eqref{eq:ApproxDiscreteTimeModel4}. With this switch the convolution is now between the scaling functions and wavelets for the field and the scaling functions and wavelets for the connectivity kernel. This convolution can be calculated analytically when using B-spline scaling and wavelet functions and can be predefined, since these are static functions. This has major computational advantages when solving the estimation problem. %\parham{Although this switch holds theoretically it should be noted that it will introduce numerical error, specially close to the edges, due to the finite extent of the field. } 
The predefined convolution can be written as
\begin{equation}\label{eq:Gammaij}
	\left[\Gamma(r')\right]_{:i} = T_s \mathbf{\Lambda}_{x}^{-1}\int_{\Omega} \boldsymbol\mu\left(r\right)\boldsymbol\lambda\left(2c_{i} + r'-r\right) \,\mathrm{d}r,
 where $\left[\Gamma(r')\right]_{:i}$ denotes the $i^{th}$ column of  $\left[\Gamma(r')\right]$ giving
	\int_{\Omega} \Gamma\left(r'\right)f\left(\boldsymbol\mu^\top\left(r'\right) \mathbf{x}_t\right) 
	\, \mathrm{d}r' \boldsymbol\theta
	+ \mathbf w_t.
% Substituting Eq.~\eqref{eq:KernelFiniteExpansion} and Eq.~\eqref{eq:FieldFiniteExpansion} in Eq.~\eqref{eq:DiscreteTimeModel}, pre-multiplying by $\boldsymbol\mu\left(r\right)$, and integrating over the space gives
% \begin{align}\label{eq:DecomposedModel2} 
% 	&\mathbf{\Lambda}_{x} \mathbf{x}_{t+1} = 
% 	\xi\mathbf{\Lambda}_{x} \mathbf{x}_t +T_s \varsigma \mathbf{\Lambda}_{\theta}\mathbf{x}_t +\int_{\Omega}\boldsymbol\mu\left(r\right)e_t\left(r\right)dr.
% \end{align}
% Cross-multiplying by $\mathbf{\Lambda}_{x}^{-1}$ gives the state transition equation
% \begin{align}\label{eq:StateEquation}
%  \mathbf x_{t+1} &=\mathbf A(\boldsymbol \theta) \mathbf x_t+ \mathbf w_t\\
% \label{eq:A_theta}
%  \mathbf A(\boldsymbol \theta) &= T_s\varsigma\mathbf{\Lambda}_{x}^{-1}\mathbf{\Lambda}_{\theta}+\xi\mathbf I.
% \end{align}

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{setspace} 
% \usepackage[usenames,dvipsnames]{color}  
%This bit is how to use hyperlink without  coloured rectangles around it
\usepackage{float}
% \usepackage{hyperref}
% \hypersetup{colorlinks,linkcolor=black,filecolor=black,urlcolor=black,citecolor=black}
% \newcommand{\dean}[1]{\textcolor{green}{#1}}
% \newcommand{\parham}[1]{\textcolor{blue}{#1}}
% \newcommand{\vik}[1]{\textcolor{red}{#1}}

\journal{Neuroimage}
\begin{document}
\begin{frontmatter}
\title{Spatiotemporal Multi-Resolution Approximation of the Amari Type Neural Field Model}

@@

\begin{abstract}
%\boldmath
Neural fields are spatially continuous state variables described by integro-differential equations, which are well suited to describe the spatiotemporal evolution of cortical activations on multiple scales. Here we develop a multi-resolution approximation (MRA) framework for the integro-difference equation (IDE) neural field model based on semi-orthogonal cardinal B-spline wavelets. In this way, a flexible framework is created, whereby both macroscopic and microscopic behavior of the system can be represented simultaneously. State and parameter estimation is performed using the Expectation Maximization (EM) algorithm. A synthetic example is provided
to demonstrate the framework.
\end{abstract} 
\begin{keyword}
  Neural Field Model, multi-resolution approximation (MRA), Expectation Maximization (EM) algorithm
\end{keyword}
\end{frontmatter}
\section{Introduction}
The human cerebral cortex has a multi-resolution architecture, where spatial scales for information processing range from ion channels, to single neurons, to networks of millions of neurons. This multi-resolution cortical architecture poses major modeling challenges to efficiently describe the brain's dynamics. This paper introduces a multi-resolution data-driven framework for neural field modeling, the Multi-Resolution Approximate Integro-Difference Equation (MRAIDE) to address this challenge.

@@

The utility of these models for answering questions in clinical neuroscience and neurology lies in the ability to infer physiological parameters from electrophysiological data. The ability to create patient-specific neural models will contribute to our knowledge base of diseases such as epilepsy and will enable the development of new treatment strategies. Large-scale neuroinformatic efforts, such as the Virtual Brain project (www.thevirtualbrain.org \citep{Jirsa2010}), are on the way to enable patient-specific full brain modeling. This is particularly relevant to the advent of new devices that use therapeutic electrical stimulation. Stimulation strategies for devices currently operate in an open loop, where stimulation parameters are chosen using a process of trial and error. Therefore, there exists enormous potential to improve the performance of these devices using systems theory, which in turn requires a suitable dynamic model. Neural mass and neural field models are ideal candidates for estimation algorithms due to their strong links with physiology and their parsimony. It is expected that parameters of the neural fields models will be patient-specific and will, therefore, need to be inferred from data.

@@

\section{Multi-Resolution Cortical Architecture}
The human neocortex has an inherent multi-resolution architecture. The spatiotemporal neural field can be written as 
\begin{equation}
	v\left(\mathbf{r},t\right) = \sum_{j} v_j\left(\mathbf{r},t\right),
\end{equation}
where $v\left(\mathbf{r},t\right)$ is a voltage field that may be measured with a multi-electrode array or voltage sensitive dye, $\mathbf{r}\in \mathbb{R}^n$ (where $n\le3$) describes the spatial location within the field, $t$ denotes time, and $j$ indexes the different levels of resolution in the field. This multi-resolution structure is depicted in Fig.~\ref{fig:MultiLayerFieldModel}.

@@

Typically, the experimentalist only has access to the net field generated by the multi-resolution architecture of the cortex. Following this, it is natural to ask how do we best describe the multi-resolution structure of the neural system. The solution to this problem was proposed by \citet{Breakspear2005} in using a wavelet decomposition. This paper is directed at developing a framework for patient-specific multi-resolution neural field models, forming a fundamental link between data and theory. 
\section{IDE Neural Field Model}
In a small network with no time delays the single layer homogeneous neural field model, in the most general form, is described by the equation
\begin{equation}
	\label{FullDoubleIntModel} \underbrace{v\left(\mathbf{r},t\right)}_{\text{field}} =
	\int_{-\infty}^t 
	\underbrace{h\left(t - t'\right)}_{\text{synapse}} \int_\Omega
	\theta\underbrace{w\left(\mathbf{r},\mathbf{r}'\right)}_{\text{connectivity}} 
	\underbrace{f\left( v\left( \mathbf{r}',t'\right)\right)}_{\text{firing}}
	\, \mathrm{d}\mathbf{r}'\mathrm{d}t',
\end{equation}
where the post-synaptic membrane voltage at time $t$ of a population of neurons at position $\mathbf r$ within the spatial domain $\Omega$ is denoted by $v\left(\mathbf r,t\right)$. The function $h(t)$ describes the post-synaptic response kernel, $w\left(\mathbf{r},\mathbf{r}'\right)$ describes the connectivity kernel and $f(v(\mathbf{r},t))$ describes the relationship between the mean membrane potential and the mean firing rate. The parameter $\theta$ can be thought of as the synaptic efficiency or coupling strength, where $\theta>0$ and $\theta<0$ describe excitation and inhibition, respectively. There are various forms of $h(t)$ that have been proposed in the literature, with different levels of complexity. These are shown in Fig.~\ref{fig:Modelcomponents}. Similarly, both the connectivity kernel and the activity function (firing rate) can take various forms. These are also shown in Fig.~\ref{fig:Modelcomponents}. The form of $w(\mathbf{r},\mathbf{r}')$ typically describes a Laplacian (exponential) or Gaussian decay in connectivity strength from any point in the field \cite{Breakspear2010,Atay2005}. By assuming the form of $h(t)$ is the same across layers (scales) the Amari style neural field model is formed, incorporating a multilayer structure into the connectivity kernel. In this way, the connectivity kernel typically takes the form of a Mexican hat (Gaussian shape within layer connectivity) or Wizard hat (Laplacian shape within layer) as depicted in Fig.~\ref{fig:Modelcomponents}.

The stochastic IDE form of the Amari neural field formulation~\cite{Amari1977} is given by (see~\cite{Freestone2011} for a full derivation)
\begin{equation}\label{eq:DiscreteTimeModel}
	v_{t+1}\left(\mathbf{r}\right) = 
	\xi v_t\left(\mathbf{r}\right) + 
	T_s \int_\Omega { 
	    w\left(\mathbf{r},\mathbf{r'}\right)
	    f\left(v_t\left(\mathbf{r}'\right)\right) 
	\, \mathrm{d}\mathbf{r}'} 
	+ e_t\left(\mathbf{r}\right), 
\end{equation}
where in this first-order (in time) formulation of the model the membrane dynamics are included through the parameter $\xi=1-T_s/\tau$, $\tau$ is the membrane time constant and $T_s$ is the sampling time step. The term $e_t(\mathbf r)$ is a zero mean Gaussian disturbance, spatially colored but temporally white, with covariance function 
\begin{equation}
cov\left(e_{t}\left(\mathbf{r}\right),e_{t+t'}\left(\mathbf{r'}\right)\right)=\sigma_d\eta(\mathbf{r}-\mathbf{r'})\delta(t-t'),
\label{eq:FieldDisturbance}
\end{equation}
where $\eta(\mathbf{r}-\mathbf{r'})$ is a spatially homogeneous covariance function, $\sigma_d$ is disturbance variance and $\delta(\cdot)$ is the Dirac-delta function. %$ = \varsigma v_t(\mathbf{r})$ \cite{VanRotterdam1982,Murphy2009}.

The observation equation describing the intracranial electrophysiological recordings is given by
\begin{equation}\label{eq:ObservationEquation}
	y_t(\mathbf{r}_{n_y}) = \int_{\Omega} { m\left(\mathbf{r}_{n_y}-\mathbf{r}'\right) v_t\left(\mathbf{r}'\right) \, \mathrm{d}\mathbf{r}'} + \epsilon_t(\mathbf{r}_{n_y}), 
\end{equation}
where $m\left(\mathbf{r}_{n_y}-\mathbf{r}'\right)$ is the observation kernel at location $\mathbf{r}_{n_y}$ and  $\boldsymbol{\epsilon}_{t}\sim \mathcal{N}\left(\mathbf{0},\mathbf{\Sigma}_{\epsilon}\right)$  is an i.i.d. Gaussian white noise process and the covariance matrix $\mathbf{\Sigma}_{\epsilon}=\sigma^2_{\epsilon}\mathbf I_{n_y}$, where $\mathbf I$ denotes the identity matrix. The observation kernel models the pickup range of the electrical fields of the brain. The lead field parameters are not included in the observation equation as the intracranial recordings are considered here. 
\section{MRA of the IDE in State-Space}
The purpose of forming the multi-resolution approximation (MRA) of the integro-difference equation (IDE) is to form a multi-resolution finite-dimensional state-space model that best captures the spatiotemporal characteristics of the neural field. Given such a model, it is then possible to perform state and parameter estimation in an efficient and robust manner, thus allowing for subject-specific models. The goal of this section is to demonstrate how to formulate a state-space model in a canonical form such that standard algorithms may be applied to estimate the multi-resolution connectivity structure.

@@

The dynamic neural field is decomposed using
\begin{equation}
 v_t\left(r\right)=\sum_{l \in \mathbb{Z}}x_{t,j_{0},l} \underbrace{\phi_{j_0,l}\left(r\right)}_{\text{scaling functions}} + \sum_{j\geq j_0}^{\infty} \sum_{l \in \mathbb{Z}} \check{x}_{t,j,l} \underbrace{\psi_{j,l}\left(r\right)}_{\text{wavelets}},
\label{eq:FieldExpansion}
\end{equation}
where $x_{t,j_{0},l}$ and $\check{x}_{t,j,l} $ are the dynamic coefficients of the expansion, constituting the state vector at time $t$. The decomposition enables a separation of the spatial characteristics, which are considered static functions that are weighted by dynamic coefficients.

@@

@@

The state-transition equation can be simplified if the chosen family of scaling and wavelet functions are isotropic where
\begin{equation}
	\boldsymbol{\lambda}_{i}(r-r') = \boldsymbol{\lambda}(2\mathbf{c}_{i}+r'-r), 
\end{equation}
where $c_i$ is the centre of the $i^{th}$ basis function of the kernel decomposition. Following this, we can switch the order of operations with the inner product and convolution in Eq.~\eqref{eq:ApproxDiscreteTimeModel4}. With this switch the convolution is now between the scaling functions and wavelets for the field and the scaling functions and wavelets for the connectivity kernel. This convolution can be calculated analytically when using B-spline scaling and wavelet functions and can be predefined, since these are static functions. This has major computational advantages when solving the estimation problem. %\parham{Although this switch holds theoretically it should be noted that it will introduce numerical error, specially close to the edges, due to the finite extent of the field. } 
The predefined convolution can be written as
\begin{equation}\label{eq:Gammaij}
	\left[\Gamma(r')\right]_{:i} = T_s \mathbf{\Lambda}_{x}^{-1}\int_{\Omega} \boldsymbol\mu\left(r\right)\boldsymbol\lambda\left(2c_{i} + r'-r\right) \,\mathrm{d}r,
\end{equation} 
 where $\left[\Gamma(r')\right]_{:i}$ denotes the $i^{th}$ column of  $\left[\Gamma(r')\right]$ giving
\begin{equation}
	\mathbf{x}_{t+1} = 
	\xi \mathbf{x}_t + 
	\int_{\Omega} \Gamma\left(r'\right)f\left(\boldsymbol\mu^\top\left(r'\right) \mathbf{x}_t\right) 
	\, \mathrm{d}r' \boldsymbol\theta
	+ \mathbf w_t.
\end{equation}
The observation equation of the state-space model is found by substituting decomposition \eqref{eq:FieldFiniteExpansion}
 into Eq.~\eqref{eq:ObservationEquation} giving
\begin{equation}\label{eq:ReducedObservationEquation} 
	\mathbf{y}_t = \mathbf{C}\mathbf{x}_t + \boldsymbol{\epsilon}_t,
\end{equation}
where each element of the observation matrix, $\mathbf{C}$, is given by
\begin{equation}\label{eq:Observationmatrix}
	\mathbf{C}_{ik} \triangleq \int_{\Omega}m(r_i -r')\boldsymbol{\mu}_k(r') \, \mathrm{d}r'.
\end{equation}
\section{B-Spline MRAIDE in State-Space }
\subsection{A Brief Introduction to B-Spline Wavelets}
This section is included to provide the interested reader with an overview of the theory used in constructing the wavelets that form the basis of the multi-resolution model. Although a deep understanding of the theory presented in this section is not required for an understanding of the key results of the paper, it serves as a reference for implementing the methods.
